Interactive Visualization of Hierarchically Structured Data

### 2016-12-06 treelapse.Rmd

---
title: "Interactive Visualization of Hierarchically Structured Data"
author: "Kris Sankaran"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_book:
    keep_tex: yes
bibliography: ["bibliography.bib"]
link-citations: yes
---

```{r set-chunk-opts}
knitr::opts_chunk$set(fig.align = "center", out.width = "200px", echo = FALSE)
```

## outline ##

This is an outline for the treelapse article I want to write.

Introduction

- Give core theses
    + Using principles from the data visualization literature, we have designed
      a tool that we hope will be useful for analyzing tree structured data,
      especially hierarchically structured collections of time series.
    + Focus / context and linking are central to each of our visualizations
    + We hope our implementation makes this specific tool more widely
      accessible
- Describe paper structure

Problem Motivation

- Microbiome motivation
      + The meeting of ideas from medicine and ecology
      + The differential abundance problem: treatment vs. control in
        experiments
      + The ecological dynamics problem
             * Are there niches?
             * Is there competition?
- Abstraction
      + Random variables at nodes
	  + In differential abundance, we have a vector of counts, one coordinate
        per group
	  + For microbial dynamics, we have a time series at nodes
	  + In these applications, we define values at nodes as linear combinations
        of descendants (sum or average). This sort of constraint is not
		necessary though
- Problems with related structure
      + Spatial hierarchy when working with time series
      + Creating trees
             * Hierarchical clustering
             * CART
             * When variables don't have intrinsic order, but have comparable
               ranges, can interpret time series as parallel coordinates

Solution Principles / Background Literature

- Focus and Context (summary then examples)
      + Degree of interest -- a way of navigating graph structures. Modularize
        the problem of graph layout and interaction. One part is determination
        of a degree of interest for nodes. The other is determining a layout
        given a particular degree of interest distribution. Blocks are trimmed
        away depending on their average degree of interest.
      + Timeboxes -- A way of querying the types of shapes that appear in a
        given collection of time series. You draw boxes, and it highlights
        series that pass through them. The focus is on highlighted series, the
        context is the collection of series in the background.
      + Visual scent -- Determine pan and zoom based on a smaller version of
        the larger figure. Similar to interactively adjusting bounds
        numerically, but more intuitive. Focus is on the part of the scent that
        is highlighted, context is the rest of the scent.
- Linking (summary then examples)
      + Linked scatterplot brushes: Let you see many dimensions for the same
        few points. Conditional on having high values for one series, what are
        values for other variables? This is analogous to timeboxes on parallel
        coordinates plots.
      + Visual queries / conditional interpretation: Power is in querying one
        variable and seeing values in the other.
      + XGobi: Historical example of linking views, grounded in
        multidimensional scaling

- Participation in broader R software ecosystem
    + Can iterate through many interactive visualizations (like in ggvis)
    + Limited scope, but easier to maintain and more likely to remain relevant
    + Rely heavily on htmlwidgets
         * Also note similarity with networkD3: though, we have focused more on
           principles for a specific problem setup, rather than constructing
           foundational building blocks (in a way, this is more applied)

Specific Proposals

- DOI Trees
     + definition and (microbiome) use case: for a single sample, where are
       most bacteria?
     + principles it relies on: just DOI tree.
     + This is the only visualization that is completely thought through from
       before (but which as far as we can tell hasn't had easily accessible
       implementations).
     + well not exactly...we are also encoding abundance in edge widths / node
       sizes
- DOI Sankeys
     + Definition and use case: treatment vs. control in tree structured data,
       but using DOI principle
     + Exploratory analog of structSSI
     + Minor note: we use the colorbrewer palette for simpler readability
- Timebox Trees
     + Combines a few more ideas: linking and focus-plus-context
     + definition and explanation of how these ideas being implemented
              * timebox recap: we're able to highlight lines passing through
                intersection of boxes (this is focus and context)
              * Link the display between tree and time series. Interpretation:
                Position on tree conditional on time series
                shape. Alternatively, database query on one field, returning
                over all of them.
              * Focus and context / pan and zoom using visual scent [so we have
                two focus and context ideas, and one linking idea, in this one
                display]
      + Primary use case this is designed for -- we notice a pattern in how a
        clump of microbial series look like, and we try to figure out which
        subtrees have this behavior (or, if that behavior is appear at
        different places on the tree).
- Treeboxes
       + Describe as the natural analog of timebox trees
       + Intentionally avoided queries on both tree and time series
       + Example use case: we can see how different parts of the tree have
         different series.
               * Or, we can look at all the phyla / orders at once
               * Or, we can observe if there is any competition within a
                 particular clade
       + Conditional on coming from a part of a tree, what are the observed
         series?

Case Studies

- Introduction
      + Practically, want to provide example output and how to interpret it
      + This is only a partial goal (there are many vignettes that describe
        this in more detail and are explicitly designed as tutorials) -- more
        of the focus here is on explaining how the visual principles described
        above can substantively improve data analysis in this problem domain
      + Purpose is to motivate potential workflows that incorporate the
        treelapse package. Show how it can be incorporated into any existing R
        data analysis workflow (this was all done in a single R document)
      + Argue that the problem of navigating hierarchical collections of time
        series is actually more generic than you might think

- Antibiotics
      + Study background: (check out abstract to confirm) How does microbiota
        respond to a time course of antibiotics. Especially relevant given
        growing understanding of importance of microbes in *health* (not just
        disease)
      + Framing as a timebox / treebox problem
      + Initial view: can clearly discern decrease across all microbes
      + Selecting some series, notice that the Firmicutes are more strongly
        affected
      + Can zoom into period around perturbation, using the scent idea
      + Can rerun using node averaged data -- this is only possible because we
        have linked it with R (it would not be possible in a standard d3 Block,
        for example)
      + Ideally: comparison with gPCA? Show how it complements.
- DOI Sankey: Preterm births
      + Basic goal of the study -- are certain microbiome signatures associated
        with preterm births?
      + Formal inferential approach was taken by the original publication,
        testing each microbe for an association with term / preterm
      + Also had an unsupervised analysis: cluster into CSTs
                - We can follow-up this unsupervised work using treelapse (this
                  is analogous to gPCA followup)
                - They used a heatmap, which gets the main points across, but
                  without the tree context (or allowing interactive focus +
                  context)
      + We immediately note that CST 4 has much more diversity
      + navigating down the tree, we notice that the main way the other CSTs
        are different is in the different types of Lactobacillus
- Housing prices
      + A non-microbial example, but with relatively obvious hierarchical
        (spatial) structure
      + Basic goal: How were home prices affected by recession, and how do
        prices and trends in prices differ across regions?
      + Findings from timebox / treebox:
                 * Can see quickly which neighborhoods are richer than others
                 * Can also see that poorer areas were more strongly affected
                   by the recession
                 * Can also see the variation in home prices within individual
                   counties
      + Limitations: We never actually see the spatial locations for the
        series. Can easily imagine a linking between the map and the original
        series (cite xgobi), but this is out of scope (at least for now)
- Bikesharing
      + A non-microbial example, and with non-obvious hierarchical structure
      + To induce hierarchical structure, we perform classification using a
        supplementary data table
                * We take advantage of the fact that this is essentially a two
                  table problem (one whose rows are different time series,
                  another giving the supplemental data)
                * Perform classification using classification trees (similar to
                  Kaggle, but not quite -- we treat a whole series as a sample,
                  not just a single time point).
       + Observations
                * Quickly notice difference between weekend and weekday
                * Also more subtle differences though -- temperature and year
                  effects, for example
                * The 3AM people on warm weekends (importance of focus)
       + Ideal scenario: Develop model using ideas from here, relate model to
         hierarchical structure again. Alternatively, fit a tree on correct and
         incorrect cases from another algorithm, inspect what are features of
         incorrectly classified cases, build these into new features.
- Potentially: genetics and hierarchical clustering?
- Summary: Tree structured data and collections of related time series are
  common than you might suspect, the principles of data visualization actually
  facilitate navigation of high dimensional data, and since it's in R, it's
  easy to alternate between interactive visualization, data regeneration /
  preparation, and statistical modeling.

Conclusion

- Potential future work
   + Arrangement
        * Cognostics for arranging views
        * Faceting timebox / treebox views
        * Linking with DPCOA / gPCA? In practice, microbiome literature cares a
          lot about these ordinations, we can try to explicitly tie them
          together in one visualization (rather than iterating, like we do
          now).
   + Incorporating the DOI idea into the timebox trees?
   + Other layouts? Graphs? DAGs?
   + Grammar
        * Is it possible to modularize the key ideas, so we can create similar
          kinds of linked and arranged views without so much effort?
- Summary
    + Developed an R package based on ideas from the visualization community
    + Think there is power in interactive visualization for hands on data
      analysis, not just polished public tools
    + Think there is a relationship between statistical modeling and data
      visualization: it can be more useful to visualize raw data in relation to
      models
    + Focused on solving a very specific methodological problem, rather than
      building a generic system. But this methodological problem appears in a
      variety of domains.

## text ##

@@

@@

In this work, we describe methods for visualization of data structured
along trees, especially hierarchically structured collections of time
series. Influenced by Degree-of-Interest trees [@heer2004doitrees],
linked scatterplot brushing [@becker1987brushing], and timeboxes
[@hochheiser2004dynamic], we have adapted the fundamental
visualization principles of "focus-plus-context" and "linking" to study
questions that arise when analyzing tree structured data. Our
motivating application is to the analysis of microbial time series,
where an evolutionary tree relating microbes is available a
priori. However, we have identified common problem types where, if a
tree is not directly available, it can be constructed from data and
then studied using our techniques.

@@

The focus-plus-context principle is that large collections of visual
elements can be studied at multiple scales, by simultaneously
focusing" on a few elements of interest and maintaining the "context"
of the coarser-scale background. A simple example of this idea is to
include a search box that highlights matching samples (focus) and
mutes the rest (context). Two more sophisticated methods anchored in
this idea are timeboxes and Degree-of-Interest (DOI) trees; both are
central to the proposals in treelapse [@hochheiser2004dynamic,
heer2004doitrees]. In timeboxes, a collection of time series are
graphically queried using interactive brushes. Series that pass
through all of the user-specified brushes are highlighted, and the
rest are faded to the background. Hence, time series meeting the
constraints imposed by the brushes are focused, while the remainder
are de-emphasized, though they remain present as context. This method
can be interpreted programmatically as the visual analog of a database
query, or probabilistically as the conditional distribution for the
full series, given it passes through certain bounds.

@@

Finally, we can study the behavior of the regression tree itself using
the DOI sankey. Here, we group samples according to their quintile of
8AM demand and then count the abundance of the groups flowing down
different branches. We find that the quintiles are each rather
strongly separated after descending even a few steps down the
regression tree – for example, Figures \@ref(fig:weekday_2011) and
\@ref(fig:weekday_2012) focuse on 2011 vs. 2012 split among weekday
samples, showing that this split distinguishes between samples falling
in the second and third quintiles of 8AM demand.

This interactive representation of regression trees is potentially
more useful on larger trees that cannot be easily parsed in a single
view; in this sense the bikesharing tree is relatively
simple. Further, in light of research on approximating more complex
supervised learning algorithms using decision trees (Boz 2002, Saito
and Nakano (2002)), this approach of modeling and interpreting
two-table data could be used in conjunction with more sophisticated
regression methods. In our ideal data analysis workflow, we imagine
the analyst applying interactive visualization and modeling techniques
in an iterative, nonlinear fashion.

@@

```{r weekday_2011, fig.cap = "Weeday demand appears larger in 2012 than 2011 -- compare with the next figure."}
knitr::include_graphics("weekday_2011.png")
```

```{r weekday_2012, fig.cap = "Weekday demand increased in 2012 -- compare with the previous figure."}
knitr::include_graphics("weekday_2012.png")
```

```{r warm_weekend, fig.cap = "The samples with highest night demand tend to fall on warm weekends."}
knitr::include_graphics("warm_weekend.png")
```

```{r bike_sankey, fig.cap = "We can interpret the regression tree using an interactive DOI representation."}
knitr::include_graphics("bike_sankey.png")
```

@@

This package has only developed basic ideas, and there are a number of
potentially useful extensions worth exploring. For example, we have
not considered the principle of arrangement in our visualizations
(Buja, Cook, and Swayne 1996), though many of our conclusions were
based on comparing alternative selections of the same display. We
could imagine faceting our displays across groups to make these types
of comparisons more accessible. Further, we have only worked with the
DOI distribution described in (Heer and Card 2004). It would be
interesting to define a more statistical notion of interest along
nodes, based on cognostics, for example (Hafen et al. 2013,
J. H. Friedman and Stuetzle (2002)). A simple extension could be to
allow graph layouts instead of trees in time and treebox displays, for
data that cannot be coerced into a hierarchical structure. Finally, if
these ideas turn out to be useful in practice, it would be valuable to
modularize the basic visualization layouts and relationships into a
library, allowing the wider community to construct novel linked,
interactive graphics with minimal effort.

In summary, we have built an easily accessible R package for
visualization techniques in a very specific methodology problem –
analysis of differential abundance and dynamics in hierarchically
structured data – that appears in a variety of application domains. We
have leveraged (Vaidyanathan et al.  2014) to create exploratory
visualizations while still deep in data analysis; in this way our work
is a departure from the culture of polished, journalistic
visualizations prioritized by the D3 community. Finally, we have given
a series of examples to demonstrate how the general visualization
techniques of focus-plus-context and linked brushing can be
practically incorporated into a range of practical analysis workflows,
from studying the impact of microbes on human health to better
allocating units in commuter bikesharing systems.

>>>>

---
title: "Interactive Visualization of Hierarchically Structured Data"
author: "Kris Sankaran and Susan Holmes"
date: "`r Sys.Date()`"
output:
    bookdown::pdf_book:
        toc: false
bibliography: ["bibliography.bib"]
biblio-style: "agsm"
link-citations: yes
---

@@

```{r set_chunk_opts, echo = FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  out.width = "250px",
  echo = FALSE,
  cache = TRUE
)
```

@@

In this work, we describe methods for visualization of data structured
along trees, especially hierarchically structured collections of time
series. Influenced by Degree-of-Interest trees [@heer2004doitrees],
linked scatterplot brushing [@becker1987brushing], and timeboxes
[@hochheiser2004dynamic], we have adapted the fundamental
visualization principles of focus-plus-context and linking to study
questions that arise when analyzing tree structured data. Our
motivating application is to the analysis of microbial time series,
where an evolutionary tree relating microbes is available a
priori. However, we have identified common problem types where, if a
tree is not directly available, it can be constructed from data and
then studied using our techniques.

@@

The focus-plus-context principle is that large collections of visual
elements can be studied at multiple scales, by simultaneously
focusing" on a few elements of interest and maintaining the "context"
of the coarser-scale background. A simple example of this idea is to
include a search box that highlights matching samples (focus) and
mutes the rest (context). Two more sophisticated methods anchored in
this idea are timeboxes and Degree-of-Interest (DOI) trees; both are
central to the proposals in treelapse [@hochheiser2004dynamic,
@heer2004doitrees]. In timeboxes, a collection of time series are
graphically queried using interactive brushes. Series that pass
through all of the user-specified brushes are highlighted, and the
rest are faded to the background. Hence, time series meeting the
constraints imposed by the brushes are focused, while the remainder
are de-emphasized, though they remain present as context. This method
can be interpreted programmatically as the visual analog of a database
query, or probabilistically as the conditional distribution for the
full series, given it passes through certain bounds.

@@

Finally, we can study the behavior of the regression tree itself using
the DOI sankey. Here, we group samples according to their quintile of
8AM demand and then count the abundance of the groups flowing down
different branches. We find that the quintiles are each rather
strongly separated after descending even a few steps down the
regression tree – for example, Figures \@ref(fig:weekday2011) and
\@ref(fig:weekday2012) focuse on 2011 vs. 2012 split among weekday
samples, showing that this split distinguishes between samples falling
in the second and third quintiles of 8AM demand.

This interactive representation of regression trees is potentially
more useful on larger trees that cannot be easily parsed in a single
view; in this sense the bikesharing tree is relatively simple. In our
ideal data analysis workflow, we imagine the analyst applying
interactive visualization and modeling techniques in an iterative,
nonlinear fashion.

@@

```{r weekday2011, fig.cap = "Weeday demand appears larger in 2012 than 2011 -- compare with the next figure."}
knitr::include_graphics("weekday_2011.png")
```

```{r weekday2012, fig.cap = "Weekday demand increased in 2012 -- compare with the previous figure."}
knitr::include_graphics("weekday_2012.png")
```

```{r warmweekend, fig.cap = "The samples with highest night demand tend to fall on warm weekends."}
knitr::include_graphics("warm_weekend.png")
```

```{r bikesankey, fig.cap = "We can interpret the regression tree using an interactive DOI representation."}
knitr::include_graphics("bike_sankey.png")
```

@@

This package has only developed basic ideas, and there are a number of
potentially useful extensions worth exploring. For example, we have
not considered the principle of arrangement in our visualizations
[@buja1996interactive], though many of our conclusions were based on
comparing alternative selections of the same display. We could imagine
faceting our displays across groups to make these types of comparisons
more accessible. Further, we have only worked with the DOI
distribution described in [@heer2004doitrees]. It would be interesting
to define a more statistical notion of interest along nodes, based on
cognostics, for example [@hafen2013trelliscope, @friedman2002john]. A
simple extension could be to allow graph layouts instead of trees in
time and treebox displays, for data that cannot be coerced into a
hierarchical structure. Finally, if these ideas turn out to be useful
in practice, it would be valuable to modularize the basic
visualization layouts and relationships into a library, allowing the
wider community to construct novel linked, interactive graphics with
minimal effort.

In summary, we have built an easily accessible R package for
visualization techniques in a very specific methodology problem –
analysis of differential abundance and dynamics in hierarchically
structured data – that appears in a variety of application domains. We
have leveraged a link between R and D3 [@vaidyanathan2014htmlwidgets]
to create exploratory visualizations while still deep in data
analysis; in this way our work is a departure from the culture of
polished, journalistic visualizations prioritized by the D3
community. Finally, we have given a series of examples to demonstrate
how the general visualization techniques of focus-plus-context and
linked brushing can be practically incorporated into a range of
practical analysis workflows, from studying the impact of microbes on
human health to better allocating units in commuter bikesharing
systems.
